---
title: "Data Science in R — Cheat Sheet by Topic"
author: "Compiled from Course Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: true
    toc_depth: 4
---

#1. Basic R — objects, classes, vectors, data frames, importing
#2. dplyr — select, filter, pipes, slice, distinct, mutate, summarize, group_by, count
#3. Tidy Data — pivot_longer, pivot_wider, separate, unite
#4. NA Handling — naniar functions
#5. Joins & Dates — all join types + lubridate
#6. Plots — boxplot, scatterplot, barplot, line, histogram, density + all aesthetics, themes, colors, faceting
#7. Maps — ggmap static + leaflet interactive
#8. Shiny Apps — selectInput, radioButtons, shinydashboard with and without sidebar

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, message=FALSE, warning=FALSE}
# Load all libraries used throughout this cheat sheet
library(tidyverse)
library(janitor)
library(naniar)
library(palmerpenguins)
library(lubridate)
library(RColorBrewer)
library(paletteer)
library(scales)
options(scipen = 999) # turns off scientific notation for the whole session
```

---

# 1. Basic R

## 1.1 Objects

```{r}
# Assign a value to an object using <- (read as "gets")
# Shortcut on Mac: Option + -
x <- 42
y <- 30

# Print object to screen by typing its name
x

# Do math with objects
x + y
x / y

# Store the result in a new object
my_experiment <- x + y
my_experiment

# Use sum() function — same result
my_experiment <- sum(x, y)
my_experiment
```

## 1.2 Data Classes

```{r}
# The four main data classes
my_numeric   <- 42
my_integer   <- 2L          # L makes it an integer
my_character <- "universe"
my_logical   <- FALSE

# Check class of an object
class(my_numeric)
class(my_integer)

# Test if something IS a specific class
is.integer(my_numeric)   # FALSE — 42 is numeric not integer

# Convert between classes using as.___()
my_numeric <- as.integer(my_numeric)
is.integer(my_numeric)   # now TRUE
```

## 1.3 Vectors

```{r}
# Create vectors using c() — stands for concatenate
my_vector    <- c(10, 20, 30)
days         <- c("Monday", "Tuesday", "Wednesday", "Thursday",
                  "Friday", "Saturday", "Sunday")
my_sequence  <- c(1:100)   # quick way to make a sequence

# Pull out elements using [] and position number
days[4]           # "Thursday"
my_sequence[10]   # 10

# Filter a vector using a logical condition
my_sequence[my_sequence <= 10]   # returns only values 1-10
```

## 1.4 Data Frames and Tibbles

```{r}
# Build vectors first, then combine into a tibble
Sex    <- c("male", "female", "male")
Length <- c(3.2, 3.7, 3.4)
Weight <- c(2.9, 4.0, 3.1)

# tibble() is the tidyverse version of a data frame
hbirds <- tibble(sex = Sex, length = Length, weight_g = Weight)
hbirds

# Inspect a data frame
names(hbirds)    # column names
dim(hbirds)      # rows x columns
str(hbirds)      # structure and types
glimpse(hbirds)  # tidyverse version of str()
summary(hbirds)  # summary statistics

# Access rows and columns using [row, col]
hbirds[1, ]    # first row
hbirds[ , 3]   # third column

# Access a column as a vector using $
mean(hbirds$length)
```

## 1.5 Importing and Exporting Data

```{r, eval=FALSE}
# Check working directory before importing
getwd()

# Read a CSV file using tidyverse read_csv()
fish <- read_csv("data/Gaeta_etal_CLC_data.csv")

# Import and immediately clean column names (lowercase + underscores)
mammals <- read_csv("data/mammal_lifehistories_v2.csv") %>% clean_names()

# Import and replace -999 with NA at load time
life_history <- read_csv("data/mammal_lifehistories_v3.csv", na = "-999") %>%
  clean_names()

# Import with multiple NA representations
sharks <- read_csv("data/SharkIncidents_1950_2022_220302.csv",
                   na = c("NA", "N/A", "Unknown", "NOT COUNTED", "", " ")) %>%
  clean_names()

# Write data to CSV — row.names=FALSE avoids printing row numbers
write_csv(hbirds, "hbirds_data.csv")
```

---

# 2. dplyr — Data Wrangling

## 2.1 select() — Choose Columns

```{r}
# Select specific columns by name
select(penguins, species, body_mass_g)

# Select a range of columns
select(penguins, species:flipper_length_mm)

# Exclude a column using !
select(penguins, !body_mass_g)

# Exclude multiple columns
select(penguins, !c(species, island, year, sex))

# Select columns containing a string
select(penguins, contains("mm"))

# Select columns starting with a string
select(penguins, starts_with("bill"))

# Select only numeric columns
select(penguins, where(is.numeric))

# Rename while selecting — new_name = old_name
select(penguins, species, body_mass = body_mass_g)

# rename() — rename without dropping other columns
rename(penguins, body_mass = body_mass_g)

# relocate() — move a column to a different position (default: first)
relocate(penguins, year)
```

## 2.2 filter() — Filter Rows

```{r}
# Filter for exact match — use == for equality
filter(penguins, species == "Adelie")

# Comparison operators: >, >=, <, <=, != (not equal)
filter(penguins, body_mass_g >= 5500)
filter(penguins, species != "Chinstrap")

# Filter using %in% — value must be one of a list
filter(penguins, species %in% c("Adelie", "Gentoo"))

# between() — inclusive range
filter(penguins, between(body_mass_g, 5250, 5500))

# near() — within a tolerance of a target value
filter(penguins, near(body_mass_g, 5400, tol = 100))

# AND: both conditions must be TRUE (use & or comma)
filter(penguins, species == "Chinstrap" & body_mass_g > 4500)

# OR: at least one condition must be TRUE (use |)
filter(penguins, species == "Chinstrap" | body_mass_g > 4500)

# Combining AND + OR — use parentheses to be explicit
filter(penguins, body_mass_g > 5000 & (bill_length_mm > 50 | bill_depth_mm > 18))
```

## 2.3 Pipes %>%

```{r}
# Pipes feed output from one function into the next
# Think of them as "and then do this"

# Select species and body_mass, THEN filter for Gentoo only
penguins %>%
  select(species, body_mass_g) %>%
  filter(species == "Gentoo")

# Select, filter, THEN sort ascending
penguins %>%
  select(species, body_mass_g) %>%
  filter(species == "Gentoo") %>%
  arrange(body_mass_g)

# Sort descending using desc()
penguins %>%
  select(species, body_mass_g) %>%
  filter(species == "Gentoo") %>%
  arrange(desc(body_mass_g))
```

## 2.4 slice() — Extract Specific Rows

```{r}
# Get the row with the smallest value
penguins %>%
  filter(species == "Gentoo") %>%
  slice_min(body_mass_g, n = 1)

# Get the top 10 rows after sorting
penguins %>%
  arrange(desc(body_mass_g)) %>%
  slice_head(n = 10)

# Get the last row
penguins %>%
  arrange(desc(body_mass_g)) %>%
  slice_tail(n = 1)
```

## 2.5 distinct() — Unique Rows

```{r}
# Check for duplicated rows across all columns
penguins %>%
  distinct()

# Find unique values in one column
penguins %>%
  distinct(species)

# Find unique combinations, keep all other columns
penguins %>%
  distinct(species, island, .keep_all = T)
```

## 2.6 mutate() — Create New Columns

```{r}
# Create a new column by transforming an existing one
penguins %>%
  mutate(body_mass_kg = body_mass_g / 1000) %>%
  select(species, body_mass_g, body_mass_kg)

# Apply a function to multiple columns using across()
# ~ starts a lambda function, . represents the current column
penguins %>%
  mutate(across(ends_with("mm"), ~ . / 10))

# Convert specific columns to a different type
penguins %>%
  mutate(across(c("species", "island"), as.factor))

# Convert all character columns to lowercase
penguins %>%
  mutate(across(where(is.character), tolower))

# if_else() — conditional transformation
# if_else(condition, value_if_TRUE, value_if_FALSE)
penguins %>%
  mutate(size = ifelse(body_mass_g > 4000, "large", "small"))

# na_if() — replace a specific value with NA
penguins %>%
  mutate(bill_length_mm = na_if(bill_length_mm, 0))

# case_when() — multiple conditions creating a new category
penguins %>%
  mutate(mass_category = case_when(
    body_mass_g < 3500               ~ "small",
    body_mass_g >= 3500 & body_mass_g < 4500 ~ "medium",
    body_mass_g >= 4500              ~ "large"
  ))
```

## 2.7 summarize() — Summary Statistics

```{r}
# Basic summary
msleep %>%
  filter(bodywt > 150) %>%
  summarize(mean_bodywt = mean(bodywt),
            min_bodywt  = min(bodywt),
            max_bodywt  = max(bodywt),
            sd_bodywt   = sd(bodywt),
            total       = n())   # n() counts rows

# Summarize all numeric columns at once
msleep %>%
  summarize(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))

# n_distinct() — count unique values
msleep %>%
  summarize(n_genera = n_distinct(genus))

# pull() — extract a single column as a vector
msleep %>%
  filter(bodywt > 150) %>%
  distinct(genus) %>%
  pull(genus)
```

## 2.8 group_by() — Grouped Summaries

```{r}
# group_by() + summarize() gives stats per group
msleep %>%
  group_by(vore) %>%
  summarize(min  = min(bodywt,  na.rm = T),
            max  = max(bodywt,  na.rm = T),
            mean = mean(bodywt, na.rm = T),
            n    = n(),
            .groups = 'keep')  # suppresses grouping warning

# Group by multiple variables
msleep %>%
  group_by(vore, order) %>%
  summarize(mean_sleep = mean(sleep_total, na.rm = T),
            n          = n(),
            .groups    = 'keep')
```

## 2.9 count() — Quick Frequency Counts

```{r}
# Count observations per category
msleep %>%
  count(vore)

# Count and sort descending
msleep %>%
  count(vore, sort = T)

# Count combinations of two variables
msleep %>%
  count(vore, order)

# tabyl() from janitor — adds percentage column
msleep %>%
  tabyl(vore)
```

---

# 3. Tidy Data — tidyr

## 3.1 pivot_longer() — Wide to Long

```{r}
# pivot_longer(cols, names_to, values_to)
# cols      = which columns to pivot
# names_to  = name of new column holding old column names
# values_to = name of new column holding old cell values

# Example: all columns except patient become rows
heartrate <- read_csv("data/heartrate.csv")

heartrate_long <- heartrate %>%
  pivot_longer(-patient,              # don't pivot the patient column
               names_to  = "drug",    # old column names go here
               values_to = "heartrate") # old cell values go here
heartrate_long
```

```{r}
# Pivot a range of columns, drop NA rows automatically
billboard <- read_csv("data/billboard.csv")

billboard %>%
  pivot_longer(wk1:wk76,
               names_to       = "week",
               values_to      = "rank",
               values_drop_na = T)   # removes rows where rank is NA
```

```{r}
# Pivot all except anchor columns
billboard %>%
  pivot_longer(-c(artist, track, date.entered),
               names_to       = "week",
               values_to      = "rank",
               values_drop_na = T)
```

```{r}
# Remove a prefix from the new column values using names_prefix
billboard %>%
  pivot_longer(
    cols           = starts_with("wk"),
    names_to       = "week",
    names_prefix   = "wk",   # removes "wk" from the values
    values_to      = "rank",
    values_drop_na = T
  )
```

## 3.2 pivot_wider() — Long to Wide

```{r}
# pivot_wider(names_from, values_from)
# names_from  = column whose values become new column names
# values_from = column whose values fill the new columns

tb_data <- read_csv("data/tb_data.csv")

tb_data %>%
  pivot_wider(names_from  = "key",
              values_from = "value")
```

## 3.3 separate() and unite()

```{r}
# separate() splits one column into two
# sep = the character to split on

heartrate2 <- read_csv("data/heartrate2.csv")

heartrate2 %>%
  separate(patient,
           into = c("patient", "gender"),
           sep  = "_")   # splits on underscore

# Split a date column into day, month, year
sydneybeaches_long %>%
  separate(date,
           into = c("day", "month", "year"),
           sep  = "/")
```

```{r}
# unite() is the opposite of separate()
# unite(new_column_name, col1, col2, sep = "separator")

heartrate3 %>%
  unite(patient_gender, "patient", "gender", sep = "_")
```

---

# 4. Handling NAs — naniar

```{r, eval=FALSE}
# miss_var_summary() — count NAs per column with percentages
miss_var_summary(life_history)

# Visualize NAs per variable as a bar chart
life_history %>%
  gg_miss_var()

# Visualize where NAs appear in a scatter plot
life_history %>%
  ggplot(aes(x = log10(mass), y = log10(wean_mass))) +
  geom_point() +
  geom_miss_point()   # highlights NA locations in a different color

# Group-level NA summary
life_history %>%
  group_by(order) %>%
  miss_var_summary(order = T)

# replace_with_na() — precise replacement per column
life_history %>%
  replace_with_na(replace = list(newborn     = "not measured",
                                 weaning     = -999,
                                 wean_mass   = -999,
                                 afr         = -999,
                                 max_life    = 0,
                                 litter_size = -999,
                                 gestation   = -999,
                                 mass        = -999))

# replace_with_na_all() — replace one value across entire data frame
life_history %>%
  replace_with_na_all(condition = ~ .x == -999)

# Use common built-in NA representations
common_na_strings   # built-in list of string NA representations
common_na_numbers   # built-in list of numeric NA representations

life_history %>%
  replace_with_na_all(condition = ~ .x %in% c(common_na_numbers, common_na_strings))
```

---

# 5. Joins and Dates

## 5.1 Joining Data Frames

```{r, eval=FALSE}
# General syntax: join_type(table1, table2, by = "key_column")

# inner_join — keeps ONLY rows with matching keys in BOTH tables
inner_join(table_A, table_B, by = "customer_ID")

# left_join — keeps ALL rows from left table, NAs where no match
left_join(table_A, table_B, by = "customer_ID")

# right_join — keeps ALL rows from right table, NAs where no match
right_join(table_A, table_B, by = "customer_ID")

# full_join — keeps ALL rows from BOTH tables, NAs where no match
full_join(table_A, table_B, by = "customer_ID")

# anti_join — rows in first table with NO match in second table
# great for troubleshooting join problems
anti_join(table_A, table_B, by = "customer_ID")

# Real example: join spider records with location data
spiders_with_locs <- left_join(spider_records, spider_locs, by = "accession")
```

## 5.2 Dates with lubridate

```{r}
# Get today's date and current time
today()
now()

# Convert strings to dates — use the function matching your date ORDER
dateformat1 <- "20200922"
dateformat2 <- "09-22-2020"
dateformat3 <- "22/09/2020"
dateformat4 <- "09-22-2020 17:00:00"

ymd(dateformat1)     # year-month-day
mdy(dateformat2)     # month-day-year
dmy(dateformat3)     # day-month-year
mdy_hms(dateformat4) # month-day-year + hours:minutes:seconds

# Convert a column in a data frame to date format
spiders_with_locs <- spiders_with_locs %>%
  mutate(date = dmy(date))
```

---

# 6. Plots — ggplot2

## The Grammar of Graphics

```{r}
# Every ggplot follows this structure:
# ggplot(data, aes(x, y)) + geom_TYPE() + other layers

# Always turn off scientific notation for cleaner plots
options(scipen = 999)
```

---

## 6.1 Boxplots — geom_boxplot()

**Use when:** comparing the range/distribution of a continuous variable across categories.

```{r}
# Basic boxplot
ggplot(data    = penguins,
       mapping = aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```

```{r}
# Boxplot with fill color by the same variable
ggplot(data    = penguins,
       mapping = aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot()
```

```{r}
# Boxplot with log scale y-axis (useful for wide-ranging data)
ggplot(data    = msleep,
       mapping = aes(x = vore, y = bodywt)) +
  scale_y_log10() +
  geom_boxplot() +
  labs(title = "Range of Bodyweight by Vore Type",
       x     = "Vore Type",
       y     = "Body Weight (log10)")
```

```{r}
# Boxplot filled by a DIFFERENT variable than x (grouped boxes)
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv",
                      na = c("", "NA", "\\")) %>%
  clean_names()

homerange %>%
  ggplot(aes(x = class, y = log10_mass, fill = taxon)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

```{r}
# Practice example: tusk length by sample year (elephants homework)
# elephants_mature %>%
#   ggplot(aes(x = years_of_sample_collection, y = tusk_length_in_cm)) +
#   geom_boxplot() +
#   labs(title = "Tusk Length for Mature Elephants",
#        x     = "Years of Sample Collection",
#        y     = "Tusk Length (cm)")
```

---

## 6.2 Scatter Plots — geom_point()

**Use when:** exploring the relationship between two continuous variables.

```{r}
# Basic scatter plot
ggplot(data    = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(na.rm = T)   # na.rm=T suppresses the NA warning
```

```{r}
# Add a linear regression line
# method=lm fits a linear model, se=T shows the confidence band
ggplot(data    = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(na.rm = T) +
  geom_smooth(method = lm, se = T)
```

```{r}
# Color points by a third variable — passed to geom_point only
# This keeps ONE regression line for all data
ggplot(data    = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(mapping = aes(color = species)) +
  geom_smooth(method = lm, se = T)
```

```{r}
# Color AND shape by species + add labels
ggplot(data    = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(mapping = aes(color = species, shape = species)) +
  geom_smooth(method = lm, se = T) +
  labs(title = "Body Mass vs. Flipper Length",
       x     = "Body Mass (g)",
       y     = "Flipper Length (mm)")
```

```{r}
# Log scale on one axis
ggplot(data    = msleep,
       mapping = aes(x = bodywt, y = sleep_total)) +
  scale_x_log10() +
  geom_point(mapping = aes(color = vore)) +
  geom_smooth(method = lm, se = T) +
  labs(title = "Body Weight vs. Total Sleep Time",
       x     = "Body Weight (log10)",
       y     = "Total Sleep Time (hrs)")
```

```{r}
# Practice example: gestation vs newborn mass (mammals data)
# mammals %>%
#   ggplot(aes(x = gestation, y = log10(newborn))) +
#   geom_point() +
#   geom_smooth(method = "lm")
```

---

## 6.3 Bar Plots — geom_bar() and geom_col()

**Use when:** counting or comparing categorical data.

```{r}
# geom_bar() — automatically counts observations
# Use when you have RAW data and want R to count for you
ggplot(data    = penguins,
       mapping = aes(x = species)) +
  geom_bar()
```

```{r}
# Fill bars with color using fill= inside aes()
ggplot(data    = penguins,
       mapping = aes(x = species)) +
  geom_bar(mapping = aes(fill = species))
```

```{r}
# Stacked bar — fill by a DIFFERENT variable than x
homerange %>%
  ggplot(aes(x = taxon, fill = trophic_guild)) +
  geom_bar() +
  coord_flip() +   # flips x and y axes — great for long category names
  labs(title = "Observations by Taxon",
       x     = "Taxonomic Group",
       y     = NULL)
```

```{r}
# Side-by-side bar — position="dodge"
homerange %>%
  ggplot(aes(x = taxon, fill = trophic_guild)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  labs(title = "Observations by Taxon",
       x     = "Taxonomic Group",
       y     = NULL)
```

```{r}
# Percentage bar — scale all bars to 100%
homerange %>%
  ggplot(aes(x = taxon, fill = trophic_guild)) +
  geom_bar(position = position_fill()) +
  scale_y_continuous(labels = scales::percent) +
  coord_flip()
```

```{r}
# geom_col() — use when you ALREADY have counts or means (x AND y specified)
life_history <- read_csv("data/mammal_lifehistories_v2.csv", na = "-999") %>%
  clean_names()

life_history %>%
  count(order, sort = T) %>%
  ggplot(aes(x = reorder(order, n), y = n)) +  # reorder() sorts bars by value
  geom_col() +
  coord_flip() +
  labs(title = "Mammal Orders by Count",
       x     = NULL,
       y     = "Count")
```

```{r}
# geom_col() with mean values + log scale
life_history %>%
  group_by(order) %>%
  summarize(mean_mass = mean(mass, na.rm = T)) %>%
  ggplot(aes(x = reorder(order, mean_mass), y = mean_mass)) +
  geom_col() +
  coord_flip() +
  scale_y_log10() +
  labs(title = "Mean Mass by Mammal Order",
       x     = NULL,
       y     = "Mean Mass (log10)")
```

---

## 6.4 Line Plots — geom_line()

**Use when:** showing change over time or a continuous sequence.

```{r}
gap <- read_csv("data/gapminder.csv") %>% clean_names()

# Basic line plot over time
gap %>%
  filter((year >= 2000 & year <= 2025) & country == "USA") %>%
  ggplot(aes(x = year, y = gdp)) +
  geom_line() +
  geom_point(shape = 5) +   # adds points at each data value
  scale_x_continuous(breaks = seq(2000, 2025, by = 5)) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "GDP per Capita in the USA, 2000-2025",
       x     = "Year",
       y     = "GDP per Capita")
```

```{r}
# Line plot with custom color and linewidth
gap %>%
  filter((year >= 2000 & year <= 2025) & country == "USA") %>%
  ggplot(aes(x = year, y = gdp)) +
  geom_line(color = "red", linewidth = 1) +
  geom_point(shape = 5) +
  scale_x_continuous(breaks = seq(2000, 2025, by = 5)) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

---

## 6.5 Histograms — geom_histogram()

**Use when:** showing the distribution of a single continuous variable.

```{r}
# Basic histogram
gap %>%
  filter(year == 1935) %>%
  ggplot(aes(x = lex)) +
  geom_histogram()
```

```{r}
# Histogram with custom color, fill, and bin count
# color = bar outline, fill = bar interior, bins = number of bars
gap %>%
  filter(year == 1935) %>%
  ggplot(aes(x = lex)) +
  geom_histogram(color = "black",
                 fill  = "turquoise4",
                 bins  = 30) +
  labs(title = "Distribution of Life Expectancy, 1935",
       x     = "Life Expectancy (years)")
```

```{r}
# Histogram + density curve on the same plot
# after_stat(density) rescales the histogram to match the density y axis
gap %>%
  filter(year == 1935) %>%
  ggplot(aes(x = lex)) +
  geom_histogram(aes(y = after_stat(density)),
                 color = "black",
                 fill  = "turquoise4") +
  geom_density(color = "red") +
  labs(title = "Distribution of Life Expectancy, 1935")
```

---

## 6.6 Density Plots — geom_density()

**Use when:** showing a smoothed distribution — like a histogram without bins.

```{r}
# Basic density plot
gap %>%
  filter(year == 1935) %>%
  ggplot(aes(x = lex)) +
  geom_density(color = "turquoise4",
               fill  = "turquoise4",
               alpha = 0.4) +   # alpha controls transparency
  labs(title = "Distribution of Life Expectancy, 1935")
```

```{r}
# Overlapping density plots — compare two years side by side
# Must convert year to factor so ggplot treats it as discrete (for coloring)
gap %>%
  filter(year %in% c(1930, 1950)) %>%
  mutate(year = factor(year)) %>%
  ggplot(aes(x = lex, color = year, fill = year)) +
  geom_density(alpha = 0.3) +
  labs(title = "Life Expectancy Distribution: 1930 vs 1950",
       x     = "Life Expectancy (years)",
       y     = NULL,
       color = "Year",
       fill  = "Year")
```

---

## 6.7 Plot Aesthetics

### Labels — labs()

```{r}
# Add/change title, axis labels, and legend titles
ggplot(data    = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(mapping = aes(color = species)) +
  labs(title = "Body Mass vs. Flipper Length",
       x     = "Body Mass (g)",
       y     = "Flipper Length (mm)",
       color = "Species")    # renames the legend title
```

### Themes

```{r}
# Built-in themes — add as a layer
p <- ggplot(data    = penguins,
            mapping = aes(x = species, y = body_mass_g)) +
  geom_boxplot()

p + theme_minimal()    # clean with no background
p + theme_classic()    # simple axes only
p + theme_bw()         # black and white
p + theme_gray()       # default gray background
p + theme_linedraw()   # black lines on white
```

```{r}
# theme() lets you customize individual elements
p +
  labs(title = "Body Mass by Species") +
  theme(
    plot.title      = element_text(size = rel(1.5), hjust = 0.5), # centered bold title
    axis.text.x     = element_text(angle = 45, hjust = 1),        # rotated x labels
    legend.position = "bottom"                                     # move legend
  )
```

### Colors — RColorBrewer and paletteer

```{r}
# View a color palette before using it
display.brewer.pal(8, "Accent")

# scale_fill_brewer() = for fill aesthetic (bars, boxes)
# scale_colour_brewer() = for color aesthetic (points, lines)
homerange %>%
  ggplot(aes(x = taxon, fill = trophic_guild)) +
  geom_bar(position = "dodge") +
  coord_flip() +
  scale_fill_brewer(palette = "Accent")
```

```{r}
# paletteer — access hundreds of color palettes
my_palette <- paletteer_d("ggsci::uniform_startrek")

# Preview the palette
barplot(rep(1, 7), axes = FALSE, col = my_palette)
```

```{r}
# Custom hex code colors
my_cols <- c("#66CDAA", "#FFCD00FF")
show_col(my_cols)   # preview in the plot panel
```

### Faceting — Multi-Panel Plots

```{r}
# facet_wrap() — ribbon of panels by one variable
homerange %>%
  ggplot(aes(x = taxon, y = log10_mass, fill = thermoregulation)) +
  geom_boxplot() +
  facet_wrap(~ trophic_guild,
             nrow     = 2,
             labeller = label_both) +  # label includes variable name
  theme(axis.text.x   = element_text(angle = 60, hjust = 1),
        legend.position = "none",
        strip.text      = element_text(size = 12, face = "bold"))
```

```{r}
# facet_grid() — rows ~ columns layout
# Use . to indicate no variable for that axis
homerange %>%
  ggplot(aes(x = taxon, y = log10_mass, fill = thermoregulation)) +
  geom_boxplot() +
  facet_grid(trophic_guild ~ thermoregulation, labeller = label_both) +
  theme(axis.text.x    = element_text(angle = 60, hjust = 1),
        legend.position = "none")
```

---

# 7. Maps

## 7.1 Static Maps — ggmap

```{r, eval=FALSE}
library(ggmap)

# Register your free Stadia Maps API key (stadiamaps.com)
register_stadiamaps("YOUR API KEY HERE", write = FALSE)

# Step 1: check the range of your coordinates
spiders %>%
  select(latitude, longitude) %>%
  summary()

# Step 2: create a bounding box — f=0.03 adds 3% padding
lat  <- c(34.67, 41.80)
long <- c(-124.1, -115.5)
bbox <- make_bbox(long, lat, f = 0.03)

# Step 3: download base map tile
# maptype options: "stamen_terrain", "stamen_toner", "stamen_watercolor"
map1 <- get_stadiamap(bbox, maptype = "stamen_terrain", zoom = 7)

# Step 4: display base map
ggmap(map1)

# Step 5: add data points on top
ggmap(map1) +
  geom_point(data  = spiders,
             aes(longitude, latitude),
             size  = 1,
             color = "red",
             alpha = 0.8) +
  labs(x     = "Longitude",
       y     = "Latitude",
       title = "CA Usofila Spider Records")
```

## 7.2 Interactive Maps — leaflet

```{r, eval=FALSE}
library(leaflet)

# Get coordinate range from data
spiders %>%
  summarize(lon_min = min(longitude, na.rm = T),
            lon_max = max(longitude, na.rm = T),
            lat_min = min(latitude,  na.rm = T),
            lat_max = max(latitude,  na.rm = T))

# Store bounding box values
lon_min <- -124.13
lon_max <- -115.49
lat_min <- 34.67
lat_max <- 41.8

# Interactive map with Stadia tiles
leaflet(spiders) %>%
  addProviderTiles(providers$Stadia.StamenTerrain) %>%
  addCircleMarkers(
    lng         = ~longitude,   # ~ means "use this column"
    lat         = ~latitude,
    radius      = 2,
    stroke      = FALSE,
    fillOpacity = 0.7
  ) %>%
  addScaleBar(position = "bottomleft") %>%
  fitBounds(lng1 = lon_min, lat1 = lat_min,
            lng2 = lon_max, lat2 = lat_max)

# Interactive map with default OpenStreetMap tiles (no API key needed)
leaflet(spiders) %>%
  addTiles() %>%
  addCircleMarkers(
    lng         = ~longitude,
    lat         = ~latitude,
    radius      = 2,
    stroke      = FALSE,
    fillOpacity = 0.7
  ) %>%
  addScaleBar(position = "bottomleft") %>%
  fitBounds(lng1 = lon_min, lat1 = lat_min,
            lng2 = lon_max, lat2 = lat_max)

# Custom marker color
leaflet(sharks_fatal) %>%
  addProviderTiles(providers$Stadia.StamenTerrain) %>%
  addCircleMarkers(
    lng         = ~longitude,
    lat         = ~latitude,
    radius      = 2,
    stroke      = FALSE,
    fillColor   = "red",   # custom fill color
    fillOpacity = 1
  ) %>%
  addScaleBar(position = "bottomleft") %>%
  fitBounds(lng1 = lon_min, lat1 = lat_min,
            lng2 = lon_max, lat2 = lat_max)
```

---

# 8. Shiny Apps

## 8.1 Basic Skeleton

```{r, eval=F, echo=T}
library(shiny)

# Every shiny app has three parts: ui, server, and shinyApp()
ui <- fluidPage(
  # user interface — what the user sees and interacts with
)

server <- function(input, output, session) {
  # server — where the reactive code and outputs go
}

shinyApp(ui, server)  # combines ui and server and launches the app
```

## 8.2 selectInput() App

```{r, eval=F, echo=T}
library(shiny)

ui <- fluidPage(
  selectInput("x",                    # id used in server to access this input
              "Select X Variable",    # label shown to user
              choices  = c("bill_length_mm",
                           "bill_depth_mm",
                           "flipper_length_mm",
                           "body_mass_g"),
              selected = "bill_length_mm"),  # default selection
  selectInput("y",
              "Select Y Variable",
              choices  = c("bill_length_mm",
                           "bill_depth_mm",
                           "flipper_length_mm",
                           "body_mass_g"),
              selected = "bill_depth_mm"),
  plotOutput("plot", width = "500px", height = "400px")  # reserves space for plot
)

server <- function(input, output, session) {
  output$plot <- renderPlot({   # renderPlot() rebuilds plot whenever input changes
    ggplot(penguins,
           aes(x = .data[[input$x]],   # .data[[]] lets us use a string as column name
               y = .data[[input$y]])) +
      geom_point(na.rm = T) +
      theme_minimal()
  })
}

shinyApp(ui, server)
```

## 8.3 radioButtons() App

```{r, eval=F, echo=T}
library(shiny)

ui <- fluidPage(
  radioButtons("x",                   # id
               "Select Fill Variable", # label
               choices  = c("trophic_guild", "thermoregulation"),
               selected = "trophic_guild"),
  plotOutput("plot", width = "500px", height = "400px")
)

server <- function(input, output, session) {
  output$plot <- renderPlot({
    ggplot(data = homerange,
           aes_string(x = "locomotion", fill = input$x)) +  # aes_string() for string inputs
      geom_bar(position = "dodge", alpha = 0.8, color = "black") +
      labs(x = NULL, y = NULL, fill = "Fill Variable") +
      theme_minimal()
  })
}

shinyApp(ui, server)
```

## 8.4 shinydashboard App

```{r, eval=F, echo=T}
library(shiny)
library(shinydashboard)

# shinydashboard replaces fluidPage() with dashboardPage()
# Minimum required: dashboardHeader + dashboardSidebar + dashboardBody

ui <- dashboardPage(
  dashboardHeader(title = "My App Title"),       # top bar title
  dashboardSidebar(disable = T),                 # disable=T hides the sidebar
  dashboardBody(
    selectInput("y",
                "Select Variable",
                choices  = c("age", "height"),
                selected = "age"),
    plotOutput("plot", width = "500px", height = "400px")
  )
)

server <- function(input, output, session) {
  output$plot <- renderPlot({
    elephants %>%
      filter(!is.na(sex)) %>%
      ggplot(aes(x    = sex,
                 y    = .data[[input$y]],
                 fill = sex)) +
      geom_boxplot(alpha = 0.8, color = "black") +
      labs(title = paste("Range of", input$y, "by Sex"),
           x     = "Sex",
           y     = input$y) +
      theme_minimal()
  })
}

shinyApp(ui, server)
```

## 8.5 shinydashboard with Sidebar Inputs

```{r, eval=F, echo=T}
library(shiny)
library(shinydashboard)

# Put selectInput() inside dashboardSidebar() to move controls to the side panel
ui <- dashboardPage(
  dashboardHeader(title = "Distribution by Order"),
  dashboardSidebar(
    selectInput("x",
                "Select Order",
                choices  = sort(unique(gestation$order)),  # sort() for alphabetical
                selected = "Afrosoricida"),
    selectInput("y",
                "Select Variable",
                choices  = c("bodymass", "longevity", "gestationlength"),
                selected = "gestationlength")
  ),
  dashboardBody(
    plotOutput("plot", width = "500px", height = "400px")
  )
)

server <- function(input, output, session) {
  output$plot <- renderPlot({
    gestation %>%
      filter(input$x == .data[["order"]]) %>%
      ggplot(aes(.data[[input$y]])) +
      geom_density(fill  = "steelblue",
                   color = "black",
                   alpha = 0.6) +
      theme_light()
  })
}

shinyApp(ui, server)
```

---

*Compiled from BIS 15L / Data Science in R course notes. All code is knit-ready — Shiny chunks use eval=F, echo=T so the code displays but does not run when knitted.*
