---
title: "Data Science in R — Full Semester Cheat Sheet"
author: "Compiled from Course Notes"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: true
    toc_float: true
    toc_depth: 4
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

---

# 1. Setup & Working Directories

```{r, eval=FALSE}
# Check your current working directory
getwd()

# Set your working directory (uncomment and change to your path)
# setwd("/Users/yourname/Documents/yourfolder")

# Example: set to a subfolder
setwd("~/Desktop/BIS15L_W26_mmerridge/lab2/homework")
```

---

# 2. Objects & Basic Operations

## 2.1 Assigning Objects

```{r, eval=FALSE}
# Assign a value to object x using the gets arrow (<-)
# Shortcut on Mac: Option + -
x <- 42

# Print the object to the screen
x

# Assign more objects
y <- 30
treatment <- 36
control   <- 38

# NOTE: Do NOT use = for assignment (not conventional in R)
z = 10  # works but avoid this
```

## 2.2 Arithmetic with Objects

```{r, eval=FALSE}
# Basic math between objects
x + y
x / y

# Sum of two objects
treatment + control

# Store the result in a new object (underscore for spaces in names)
my_experiment <- treatment + control
my_experiment

# Use the sum() function — if you reuse the same object name, it is overwritten
my_experiment <- sum(treatment, control)
my_experiment

# Get help on any function
?sum
```

## 2.3 Naming Conventions

```{r, eval=FALSE}
# DO NOT name objects the same as built-in R functions
# else <- 12   # ERROR — else is a reserved word in R

# Naming an object 'mean' is allowed but confusing:
mean <- 20

# Order of operations applies; use parentheses to be explicit
2 + 8 + 2 / 3      # divides 2 by 3 first — NOT what you want
(2 + 8 + 2) / 3    # correct: 4

# Use c() (concatenate) to combine values into a vector for functions like mean()
mean(c(2, 8, 2))    # correct: 4

# This does NOT work — mean() only takes one argument for values
mean(2, 8, 2)       # wrong: only uses 2
```

---

# 3. Data Classes

```{r, eval=FALSE}
# The four main data classes in R:
my_numeric   <- 42
my_integer   <- 2L         # Adding L makes it an integer
my_character <- "universe"
my_logical   <- FALSE

# Check class
class(my_numeric)
class(my_integer)

# Test if something IS a specific class
is.integer(my_numeric)   # FALSE — 42 is numeric, not integer

# Convert to a different class using as.___()
my_numeric <- as.integer(my_numeric)  # converts numeric to integer
is.integer(my_numeric)                # now TRUE

# Convert to factor (useful for categorical variables)
hot_springs$scientist <- as.factor(hot_springs$scientist)
levels(hot_springs$scientist)   # shows the factor levels
```

---

# 4. Missing Data (NA)

```{r, eval=FALSE}
# NA is R's symbol for missing data
my_missing <- NA

# Test for NA
is.na(my_missing)    # TRUE
anyNA(my_missing)    # TRUE

# Create a vector with some NA values
new_vector <- c(7, 6.2, 5, 9, NA, 4, 9.8, 7, 3, 2)

# mean() with NA returns NA by default — NA "infects" calculations
mean(new_vector)

# Remove NA's from the calculation using na.rm = TRUE
mean(new_vector, na.rm = TRUE)

# Check how many NA's are in each column of a data frame
life_history %>%
  summarize(across(everything(), ~ sum(is.na(.))))

# Find all -999 values (a common NA placeholder in published data)
life_history %>%
  summarize(across(everything(), ~ sum(. == -999)))
```

---

# 5. Vectors

## 5.1 Creating Vectors

```{r, eval=FALSE}
# Numeric vector
my_vector <- c(10, 20, 30)

# Character vector
days_of_the_week <- c("Monday", "Tuesday", "Wednesday", "Thursday",
                      "Friday", "Saturday", "Sunday")

# Sequence vector — quick way to generate test data
my_vector_sequence <- c(1:100)
```

## 5.2 Indexing Vectors

```{r, eval=FALSE}
# Use [] to pull out a specific element by position
days_of_the_week[4]          # "Thursday" — the 4th element
my_vector_sequence[10]       # 10

# Use logical operators to test all elements
my_vector_sequence <= 10     # returns TRUE/FALSE for each element

# Use [] with a logical condition to return only matching values
my_vector_sequence[my_vector_sequence <= 10]   # returns 1 through 10
```

---

# 6. Data Frames & Tibbles

## 6.1 Building a Data Frame

```{r, eval=FALSE}
# Create individual vectors first
Sex    <- c("male", "female", "male")
Length <- c(3.2, 3.7, 3.4)
Weight <- c(2.9, 4.0, 3.1)

# Combine into a tibble (tidyverse data frame)
hbirds <- tibble(Sex, Length, Weight)

# Give custom column names at creation time
hbirds <- tibble(sex = Sex, length = Length, weight_g = Weight)

# Using base R data.frame()
mountains <- data.frame(tallest_mountains_hight_ft,
                        tallest_mountains_hight_m,
                        tallest_mountains_names)
```

## 6.2 Inspecting a Data Frame

```{r, eval=FALSE}
# Column names
names(hbirds)

# Dimensions: rows x columns
dim(hbirds)

# Structure of the data frame (shows column types)
str(hbirds)

# Similar to str(), but tidyverse style
glimpse(hbirds)

# Summary statistics for every column
summary(hbirds)

# First n rows (default 6)
head(fish)

# Last n rows (default 6)
tail(fish)

# Number of rows
nrow(fish)

# Number of columns
ncol(fish)

# Count observations in a categorical variable — fast frequency table
table(fish$lakeid)
```

## 6.3 Accessing Rows and Columns

```{r, eval=FALSE}
# Access by [row, column] — leave one blank to get all rows or all columns
hbirds[1, ]   # first row, all columns
hbirds[ , 3]  # all rows, third column

# Access a column as a vector using $
mean(hbirds$length)
```

---

# 7. Importing & Exporting Data

## 7.1 Reading Data

```{r, eval=FALSE}
# Check working directory before importing
getwd()

# Read a CSV file using tidyverse read_csv()
hot_springs <- read_csv("hsprings.csv")

# Read from a subfolder
fish <- read_csv("data/Gaeta_etal_CLC_data.csv")

# Import and immediately replace -999 with NA
life_history <- read_csv("data/mammal_lifehistories_v3.csv", na = "-999")

# Import and immediately clean column names (lowercase, underscores)
mammals <- read_csv("data/mammal_lifehistories_v2.csv") %>% clean_names()

# Import with multiple NA representations
sharks <- read_csv("data/SharkIncidents_1950_2022_220302.csv",
                   na = c("NA", "N/A", "Unknown", "NOT COUNTED", "", " ")) %>%
  clean_names()

# Import with a specific NA string and multiple NA values for a column
homerange <- read_csv("data/Tamburelloetal_HomeRangeDatabase.csv",
                      na = c("", "NA", "\\")) %>%
  clean_names()
```

## 7.2 Writing Data

```{r, eval=FALSE}
# Write a data frame to CSV — row.names = FALSE avoids adding row numbers
write.csv(hbirds, "hbirds_data.csv", row.names = FALSE)

# tidyverse version (slightly preferred)
write_csv(mountains, "mountains_data.csv")

# Save spider data with locations to CSV
write.csv(spiders_with_locs, file = "spiders_with_locs.csv", row.names = FALSE)
```

---

# 8. Cleaning Data with janitor

```{r, eval=FALSE}
library("janitor")

# clean_names() converts column names to lowercase with underscores
# Best practice: use it at import
mammals <- read_csv("data/mammal_lifehistories_v2.csv") %>% clean_names()

# You can also apply it after import
mammals <- mammals %>% clean_names()

# tabyl() produces counts and percentages for a variable (from janitor)
msleep %>%
  tabyl(vore)
```

---

# 9. dplyr — Data Wrangling

## 9.1 `select()` — Choose Columns

```{r, eval=FALSE}
library("tidyverse")

# Select specific columns by name (order matters for output)
select(penguins, species, body_mass_g)

# Select a range of columns using start:end
select(penguins, species:flipper_length_mm)

# Exclude a column using !
select(penguins, !body_mass_g)

# Exclude multiple columns using c()
select(penguins, !c(species, island, year, sex))

# Select columns that contain a string
select(penguins, contains("mm"))

# Select columns that start with a string
select(penguins, starts_with("bill"))

# Select columns that end with a string
select(penguins, ends_with("mm"))

# Select only numeric columns
select(penguins, where(is.numeric))

# Select only character columns
select(msleep, where(is.character))
```

## 9.2 `rename()` and `relocate()`

```{r, eval=FALSE}
# rename(): new_name = old_name
rename(penguins, body_mass = body_mass_g)

# Rename inside select() — selects AND renames at the same time
select(penguins, species, body_mass = body_mass_g)

# relocate() moves a column to a different position (default: first)
relocate(penguins, year)
```

## 9.3 `filter()` — Filter Rows

```{r, eval=FALSE}
# Filter for exact match (== for equality)
filter(penguins, species == "Adelie")

# Filter with comparison operators: >, >=, <, <=, != (not equal)
filter(penguins, body_mass_g >= 5500)
filter(penguins, species != "Chinstrap")

# Filter using %in% — checks if value is in a vector of options
filter(penguins, body_mass_g %in% c(5250, 5500))

# Filter with between() — inclusive range
filter(penguins, between(body_mass_g, 5250, 5500))

# Filter with near() — values within a tolerance of a target
filter(penguins, near(body_mass_g, 5400, tol = 100))

# Filter for lake "BO" in fish data
filter(fish, lakeid == "BO")

# Exclude lake "AR"
filter(fish, lakeid != "AR")

# Filter for values within 0.25 of 8
filter(fish, near(scalelength, 8, tol = 0.25))

# Filter for scalelength between 2 and 4
filter(fish, between(scalelength, 2, 4))
```

## 9.4 Multiple Conditions in `filter()`

```{r, eval=FALSE}
# AND: both conditions must be TRUE (use & or a comma)
filter(penguins, species == "Chinstrap" & body_mass_g > 4500)
filter(penguins, species == "Chinstrap", body_mass_g > 4500)  # same as above

# OR: at least one condition must be TRUE (use |)
filter(penguins, species == "Chinstrap" | body_mass_g > 4500)

# NOT condition2: condition1 TRUE but condition2 FALSE
filter(penguins, species == "Chinstrap", !body_mass_g > 4500)

# Combining AND + OR with parentheses for clarity
filter(penguins, body_mass_g > 5000 & (bill_length_mm > 50 | bill_depth_mm > 18))

# Filter for multiple lake IDs
filter(fish, lakeid == "AL" | lakeid == "AR")
# OR equivalently with %in%:
filter(fish, lakeid %in% c("AL", "AR"))
```

## 9.5 `arrange()` — Sort Rows

```{r, eval=FALSE}
# Sort in ascending order (default)
penguins %>%
  filter(species == "Gentoo") %>%
  arrange(body_mass_g)

# Sort in descending order using desc()
penguins %>%
  filter(species == "Gentoo") %>%
  arrange(desc(body_mass_g))
```

## 9.6 `slice()` — Extract Specific Rows

```{r, eval=FALSE}
# Get the row(s) with the smallest value — n= specifies how many rows
penguins %>%
  select(species, body_mass_g) %>%
  filter(species == "Gentoo") %>%
  slice_min(body_mass_g, n = 1)

# Get the top n rows (after sorting)
homerange %>%
  arrange(desc(mean.hra.m2)) %>%
  slice_head(n = 10)

# Get the last n rows
homerange %>%
  arrange(desc(mean.hra.m2)) %>%
  slice_tail(n = 1)

# See all slice options
?slice
```

## 9.7 Pipes `%>%`

```{r, eval=FALSE}
# Pipes feed output from one function into the next
# Think of them as "and then"

# Example: select species and body_mass, THEN filter for Gentoo only
penguins %>%
  select(species, body_mass_g) %>%
  filter(species == "Gentoo")

# Chain select + filter + arrange
penguins %>%
  select(species, body_mass_g) %>%
  filter(species == "Gentoo") %>%
  arrange(desc(body_mass_g))

# Complex example: fish by ID, sorted
fish %>%
  select(fish_id, radii_length_mm, scalelength) %>%
  filter(fish_id == 300) %>%
  arrange(scalelength)
```

---

# 10. dplyr — Transforming Data

## 10.1 `distinct()` — Find Unique Rows

```{r, eval=FALSE}
# Check for duplicated rows across all columns
penguins %>%
  distinct()

# Find all unique species
penguins %>%
  distinct(species)

# Find unique combinations of species + island, keep all other columns
penguins %>%
  distinct(species, island, .keep_all = T)
```

## 10.2 `mutate()` — Create New Columns

```{r, eval=FALSE}
# Create a new column by transforming an existing one
penguins %>%
  mutate(body_mass_kg = body_mass_g / 1000) %>%
  select(species, body_mass_g, body_mass_kg) %>%
  arrange(body_mass_kg)

# Convert multiple columns at once using across()
# ~ starts a lambda (anonymous function), . represents the current column
penguins %>%
  mutate(across(ends_with("mm"), ~ . / 10))

# Convert specific columns to a different type
gabon <- gabon %>%
  mutate(across(c("hunt_cat", "land_use", "transect_id"), as.factor))

# Convert all character columns to lowercase
mammals %>%
  mutate(across(where(is.character), tolower))

# Convert specific named columns to lowercase
mammals %>%
  mutate(across(c("order", "family"), tolower))

# Convert everything to lowercase — WARNING: changes numeric columns to character
mammals %>%
  mutate(across(everything(), tolower))

# Add a computed column and chain transformations
mammals %>%
  mutate(mass_kg = mass / 1000) %>%
  select(genus, species, mass, mass_kg) %>%
  arrange(desc(mass_kg))

# Convert max_life from months to years in a new column
mammals %>%
  select(genus, species, max_life) %>%
  mutate(max_life_new = max_life / 12) %>%
  arrange(desc(max_life_new))
```

## 10.3 `if_else()` and `na_if()` — Conditional Transformations

```{r, eval=FALSE}
# if_else(condition, value_if_TRUE, value_if_FALSE)
# Replace -999 with NA in a specific column
mammals %>%
  select(genus, species, newborn) %>%
  mutate(newborn_new = ifelse(newborn == -999.00, NA, newborn))

# Replace -999 across many columns
mammals <- mammals %>%
  mutate(mass        = ifelse(mass        == -999.00, NA, mass)) %>%
  mutate(gestation   = ifelse(gestation   == -999.00, NA, gestation)) %>%
  mutate(newborn     = ifelse(newborn     == -999.00, NA, newborn)) %>%
  mutate(weaning     = ifelse(weaning     == -999.00, NA, weaning)) %>%
  mutate(wean_mass   = ifelse(wean_mass   == -999.00, NA, wean_mass)) %>%
  mutate(afr         = ifelse(afr         == -999.00, NA, afr)) %>%
  mutate(max_life    = ifelse(max_life    == -999.00, NA, max_life)) %>%
  mutate(litter_size = ifelse(litter_size == -999.00, NA, litter_size)) %>%
  mutate(litters_year= ifelse(litters_year== -999.00, NA, litters_year))

# na_if() — replace a specific value with NA in a single column
life_history_no_nas <- life_history_no_nas %>%
  mutate(max_life = na_if(max_life, 0))

# Fix negative values that should be NA
amniota_tidy <- amniota_tidy %>%
  mutate(female_maturity_d = ifelse(female_maturity_d < 0, NA, female_maturity_d))
```

## 10.4 `case_when()` — Multi-Condition Transformation

```{r, eval=FALSE}
# case_when() creates a new categorical column from a numeric one
# Syntax: condition ~ "label"

income_cat2000 <- gap %>%
  filter(year == 2000) %>%
  mutate(income_category = case_when(
    gdp <= 1100                  ~ "low",
    gdp > 1100  & gdp <= 4500   ~ "low_middle",
    gdp > 4500  & gdp <= 13900  ~ "upper_middle",
    gdp > 13900                 ~ "high"
  )) %>%
  select(country, gdp, income_category) %>%
  # Convert to factor with specific level order for plotting
  mutate(income_category = factor(income_category,
                                  levels = c("low",
                                             "low_middle",
                                             "upper_middle",
                                             "high")))

# case_when() with string matching — create season categories from month
sydneybeaches_long %>%
  separate(date, into = c("day", "month", "year"), sep = "/") %>%
  mutate(season = case_when(
    month == "12" | month == "01" | month == "02" ~ "winter",
    month == "03" | month == "04" | month == "05" ~ "spring",
    month == "06" | month == "07" | month == "08" ~ "summer",
    month == "09" | month == "10" | month == "11" ~ "fall"
  ))
```

---

# 11. dplyr — Summarizing Data

## 11.1 `summarize()` — Summary Statistics

```{r, eval=FALSE}
# Basic single-variable summary
msleep %>%
  filter(bodywt > 150) %>%
  summarize(mean_bodywt_lg = mean(bodywt))

# Multiple statistics at once — all in one summarize() call
msleep %>%
  filter(bodywt > 150) %>%
  summarize(mean_bodywt_lg = mean(bodywt),
            min_bodywt_lg  = min(bodywt),
            max_bodywt_lg  = max(bodywt),
            sd_bodywt_lg   = sd(bodywt),
            total          = n())         # n() counts rows

# Summarize all numeric columns (with NA removal)
msleep %>%
  select(where(is.numeric)) %>%
  summarize_all(mean, na.rm = TRUE)

# More concise version using across()
msleep %>%
  summarize(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))

# Multiple functions across all numeric columns
elephants %>%
  summarize(
    across(
      where(is.numeric),
      list(
        mean   = ~ mean(.x,   na.rm = TRUE),
        median = ~ median(.x, na.rm = TRUE),
        sd     = ~ sd(.x,     na.rm = TRUE),
        min    = ~ min(.x,    na.rm = TRUE),
        max    = ~ max(.x,    na.rm = TRUE),
        n      = ~ sum(!is.na(.x))
      )
    )
  )
```

## 11.2 `n()`, `n_distinct()`, and `pull()`

```{r, eval=FALSE}
# n() counts total rows in a group
msleep %>%
  filter(bodywt > 150) %>%
  summarize(total = n())

# n_distinct() counts unique values
msleep %>%
  summarize(n_genera = n_distinct(genus))

# How many distinct genera have body weight > 150?
msleep %>%
  filter(bodywt > 150) %>%
  summarize(n_genera = n_distinct(genus))

# pull() extracts a single column as a vector (not a data frame)
msleep %>%
  filter(bodywt > 150) %>%
  distinct(genus) %>%
  pull(genus)
```

## 11.3 `group_by()` — Grouped Summaries

```{r, eval=FALSE}
# group_by() + summarize() = summary statistics per group
elephants %>%
  group_by(sex) %>%
  summarize(across(where(is.numeric), ~ mean(.x, na.rm = T)))

# Group by multiple variables
elephants %>%
  group_by(years_of_sample_collection, sex) %>%
  summarize(mean_estimated_age_years = mean(estimated_age_years, na.rm = T),
            n = n(),
            .groups = 'keep') %>%   # .groups = 'keep' suppresses a warning
  arrange(sex)

# Another grouped summary — min, max, mean, n by vore type
msleep %>%
  group_by(vore) %>%
  summarize(min  = min(bodywt, na.rm = T),
            max  = max(bodywt, na.rm = T),
            mean = mean(bodywt, na.rm = T),
            n    = n(),
            .groups = 'keep')

# Grouped miss_var_summary (requires naniar)
life_history_no_nas %>%
  group_by(order) %>%
  miss_var_summary(order = T)
```

## 11.4 `count()` — Quick Frequency Counts

```{r, eval=FALSE}
# Count observations per category
msleep %>%
  count(vore)

# Count and sort in descending order
msleep %>%
  count(vore, sort = T)

# Count combinations of two variables
msleep %>%
  count(vore, order)

# Count then filter for top n
msleep %>%
  filter(vore == "herbi") %>%
  count(order, sort = T)

# tabyl() from janitor — adds percentage column
msleep %>%
  tabyl(vore)
```

---

# 12. Tidy Data — `tidyr`

## 12.1 `pivot_longer()` — Wide to Long

```{r, eval=FALSE}
# Rules:
# pivot_longer(cols, names_to, values_to)
# cols       = columns to pivot
# names_to   = name of the new column that will hold old column names
# values_to  = name of the new column that will hold the values

# Example 1: all columns except patient become rows
heartrate_long <- heartrate %>%
  pivot_longer(-patient,               # don't pivot the 'patient' column
               names_to  = "drug",     # old column names become values here
               values_to = "heartrate") # cell values go here

# Example 2: pivot a range of columns, drop NAs
billboard %>%
  pivot_longer(wk1:wk76,
               names_to      = "week",
               values_to     = "rank",
               values_drop_na = T)   # automatically removes NA rows

# Example 3: pivot all columns except a set of anchor columns
billboard %>%
  pivot_longer(-c(artist, track, date.entered),
               names_to       = "week",
               values_to      = "rank",
               values_drop_na = T)

# Example 4: pivot columns matching a prefix, remove the prefix from values
billboard %>%
  pivot_longer(
    cols         = starts_with("wk"),
    names_to     = "week",
    names_prefix = "wk",              # removes "wk" from the values in the new column
    values_to    = "rank",
    values_drop_na = T
  )

# Example 5: column names encode TWO variables (experiment_replicate)
qpcr_tidy <- qpcr_untidy %>%
  pivot_longer(
    exp1_rep1:exp3_rep3,
    names_to  = c("experiment", "replicate"),
    names_sep = "_",                  # splits on underscore
    values_to = "mRNA expression"
  )
```

## 12.2 `pivot_wider()` — Long to Wide

```{r, eval=FALSE}
# Rules:
# pivot_wider(names_from, values_from)
# names_from  = column whose values become new column names
# values_from = column whose values fill the new columns

# Simple wide conversion
tb_data %>%
  pivot_wider(names_from  = "key",
              values_from = "value")

# Dates as column names
sydneybeaches_wide <- sydneybeaches_long %>%
  pivot_wider(names_from  = date,
              values_from = enterococci_cfu_100ml)

# Pivot summary data to wide for easy reading
mean_entero %>%
  pivot_wider(names_from  = site,
              values_from = mean_enterococci_cfu_100ml)

# Gene expression data — type becomes column names
gene_exp %>%
  pivot_wider(names_from  = "type",
              values_from = "L4_values")

# Pivot back to long after operations in wide format
sydneybeaches_wide %>%
  pivot_longer(-site,
               names_to  = "date",
               values_to = "enterococci_cfu_100ml")
```

## 12.3 `separate()` and `unite()`

```{r, eval=FALSE}
# separate() splits one column into multiple columns
# Syntax: separate(column, into = c("new1", "new2"), sep = "delimiter")

# Split on underscore
heartrate2 %>%
  separate(patient, into = c("patient", "gender"), sep = "_")

# Split from the end (negative position)
heartrate2 %>%
  separate(patient, into = c("patient", "gender"), sep = -1)

# Split date into day/month/year
sydneybeaches_long %>%
  separate(date, into = c("day", "month", "year"), sep = "/")

# Remove leftover separator characters using stringr
heartrate2 %>%
  separate(patient, into = c("patient", "gender"), sep = -1) %>%
  mutate(patient = str_remove(patient, "_"))

# unite() is the opposite of separate()
# unite(new_column_name, col1, col2, sep = "separator")
heartrate3 %>%
  unite(patient_gender, "patient", "gender", sep = "_")
```

---

# 13. Handling NA Values with `naniar`

```{r, eval=FALSE}
# install.packages("naniar")
library("naniar")

# miss_var_summary() — count NAs per column, with percentage
miss_var_summary(life_history_no_nas)

# Visualize NAs per variable as a bar chart
life_history_no_nas %>%
  gg_miss_var()

# Visualize where NAs appear in a scatter plot
life_history_no_nas %>%
  ggplot(aes(x = log10(mass), y = log10(wean_mass))) +
  geom_point() +
  geom_miss_point()   # highlights NA locations in a different color

# Group-level NA summary
life_history_no_nas %>%
  group_by(order) %>%
  miss_var_summary(order = T)

# replace_with_na() — precise replacement per variable
life_history %>%
  replace_with_na(replace = list(newborn     = "not measured",
                                 weaning     = -999,
                                 wean_mass   = -999,
                                 afr         = -999,
                                 max_life    = 0,
                                 litter_size = -999,
                                 gestation   = -999,
                                 mass        = -999)) %>%
  miss_var_summary()

# replace_with_na_all() — replace one value across the entire data frame
life_history %>%
  replace_with_na_all(condition = ~ .x == -999) %>%
  miss_var_summary()

# Replace using common NA strings/numbers (built-in naniar vectors)
common_na_strings   # shows the built-in list of string NA representations
common_na_numbers   # shows the built-in list of numeric NA representations

life_history %>%
  replace_with_na_all(condition = ~ .x %in% c(common_na_numbers, common_na_strings)) %>%
  mutate(newborn = na_if(newborn, "not measured"))

# Use read_csv() with na= argument for efficient NA handling at import
life_history_no_nas <- read_csv("data/mammal_lifehistories_v3.csv", na = "-999") %>%
  clean_names()

# Replace a specific value with NA after import
amniota_tidy <- amniota %>%
  replace_with_na_all(condition = ~ .x == -999)
```

---

# 14. Joining Data Frames

```{r, eval=FALSE}
library("tidyverse")

# General syntax: join_type(firstTable, secondTable, by = "key_column")

# inner_join — keeps ONLY rows with matching keys in BOTH tables
inner_join(table_A, table_B, by = "customer_ID")

# left_join — keeps ALL rows from left table, matches from right (NAs if no match)
left_join(table_A, table_B, by = "customer_ID")

# right_join — keeps ALL rows from right table, matches from left (NAs if no match)
right_join(table_A, table_B, by = "customer_ID")

# full_join — keeps ALL rows from BOTH tables (NAs where no match)
full_join(table_A, table_B, by = "customer_ID")

# anti_join — keeps rows from first table that have NO match in second table
# Useful for troubleshooting join issues
anti_join(table_A, table_B, by = "customer_ID")

# Real-world example: join spider records with location data
spiders_with_locs <- left_join(spider_records, spider_locs, by = "accession")
```

---

# 15. Working with Dates — `lubridate`

```{r, eval=FALSE}
library("lubridate")

# Get today's date
today()

# Get current date and time
now()

# Converting strings to date objects
# Use the function that matches the ORDER of your date components

dateformat1 <- "20200922"
dateformat2 <- "09-22-2020"
dateformat3 <- "22/09/2020"
dateformat4 <- "09-22-2020 17:00:00"
dateformat5 <- "20200922 170000"

ymd(dateformat1)       # year-month-day
mdy(dateformat2)       # month-day-year
dmy(dateformat3)       # day-month-year
mdy_hms(dateformat4)   # month-day-year with hours:minutes:seconds
ymd_hms(dateformat5)   # year-month-day with time

# Convert a column in a data frame to date format
spiders_with_locs <- spiders_with_locs %>%
  mutate(date = dmy(date))   # the date column was in day-month-year format
```

---

# 16. ggplot2 — Data Visualization

## 16.1 Core Syntax

```{r, eval=FALSE}
# The grammar of graphics:  plot = data + geom + aesthetics

# Minimum working plot structure:
ggplot(data = penguins,
       mapping = aes(x = species, y = body_mass_g)) +
  geom_boxplot()

# Turn off scientific notation for plots (affects the whole session)
options(scipen = 999)
```

## 16.2 Boxplots — `geom_boxplot()`

```{r, eval=FALSE}
# Basic boxplot
ggplot(data = penguins,
       mapping = aes(x = species, y = body_mass_g)) +
  geom_boxplot()

# Boxplot with fill by a categorical variable
ggplot(data = msleep,
       mapping = aes(x = vore, y = bodywt)) +
  scale_y_log10() +                          # log-scale the y axis
  geom_boxplot() +
  labs(title = "Range of Bodyweight by Vore Type")

# Boxplot comparing two categorical variables (group by taxon within class)
homerange %>%
  ggplot(aes(x = class, y = log10_mass, fill = taxon)) +
  geom_boxplot()
```

## 16.3 Scatter Plots — `geom_point()`

```{r, eval=FALSE}
# Basic scatter plot
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point()

# Remove NA warning
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(na.rm = T)

# Add a linear regression line (se = T shows confidence band)
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(na.rm = T) +
  geom_smooth(method = lm, se = T)

# Color points by a third variable (passed to geom_point only)
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(mapping = aes(color = species)) +
  geom_smooth(method = lm, se = T)   # single regression line for all data

# Color AND shape by species
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(mapping = aes(color = species, shape = species)) +
  geom_smooth(method = lm, se = T)

# Jitter (useful when points overlap)
p + geom_jitter(size = 3)

# Custom point appearance (no variable mapping — global aesthetics)
p + geom_point(size = 3, color = "black", alpha = 0.6, shape = 6, fill = "lemonchiffon4")

# See all available color names
colors()

# Log-scale axes
ggplot(data = msleep,
       mapping = aes(x = bodywt, y = sleep_total)) +
  scale_x_log10() +
  geom_point(mapping = aes(color = vore)) +
  geom_smooth(method = lm, se = T)
```

## 16.4 Bar Plots — `geom_bar()` and `geom_col()`

```{r, eval=FALSE}
# geom_bar() — counts observations in a categorical variable automatically
ggplot(data = penguins,
       mapping = aes(x = species)) +
  geom_bar()

# Fill inside bars by the same variable (adds color)
ggplot(data = penguins,
       mapping = aes(x = species)) +
  geom_bar(mapping = aes(fill = species))

# Stacked bar: fill by a DIFFERENT variable
homerange %>%
  ggplot(aes(x = taxon, fill = trophic_guild)) +
  geom_bar() +
  coord_flip()

# Side-by-side bar: position = "dodge"
homerange %>%
  ggplot(aes(x = taxon, fill = trophic_guild)) +
  geom_bar(position = "dodge") +
  coord_flip()

# Percentage bar: scale to 100%
homerange %>%
  ggplot(aes(x = taxon, fill = trophic_guild)) +
  geom_bar(position = position_fill()) +
  scale_y_continuous(labels = scales::percent) +
  coord_flip()

# geom_col() — when you have pre-counted data (x AND y specified)
life_history %>%
  count(order, sort = T) %>%
  ggplot(aes(x = order, y = n)) +
  geom_col() +
  coord_flip()

# geom_col() with mean values — reorder() sorts bars by value
life_history %>%
  group_by(order) %>%
  summarize(mean_mass = mean(mass, na.rm = T)) %>%
  ggplot(aes(x = reorder(order, mean_mass), y = mean_mass)) +
  geom_col() +
  coord_flip() +
  scale_y_log10()
```

## 16.5 Line Plots — `geom_line()`

```{r, eval=FALSE}
# Line plot over time
gap %>%
  filter((year >= 2000 & year <= 2025) & country == "USA") %>%
  ggplot(aes(x = year, y = gdp)) +
  geom_line() +
  geom_point(shape = 5) +
  scale_x_continuous(breaks = 2000:2025, minor_breaks = NULL) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  labs(title = "GDP per Capita in the USA, 2000-2025",
       x = "Year",
       y = "GDP")

# Line plot with custom color and wider spacing on x breaks
gap %>%
  filter((year >= 2000 & year <= 2025) & country == "USA") %>%
  ggplot(aes(x = year, y = gdp)) +
  geom_line(color = "red") +
  geom_point(shape = 5) +
  scale_x_continuous(breaks = seq(2000, 2025, by = 5)) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

# Line plot with custom linewidth
gapminder %>%
  filter(country == top_country) %>%
  ggplot(aes(year, pop)) +
  geom_line(color = "steelblue", linewidth = 1)
```

## 16.6 Histograms — `geom_histogram()`

```{r, eval=FALSE}
# Basic histogram
gap %>%
  filter(year == 1935) %>%
  ggplot(aes(x = lex)) +
  geom_histogram()

# Histogram with custom color (outline), fill, and number of bins
# color = bar outline, fill = bar interior
gap %>%
  filter(year == 1935) %>%
  ggplot(aes(x = lex)) +
  geom_histogram(color = "black", fill = "turquoise4", bins = 30) +
  labs(title = "Distribution of Life Expectancy")

# Combine histogram and density curve on the same plot
# after_stat(density) scales the histogram so it matches the density y axis
gap %>%
  filter(year == 1935) %>%
  ggplot(aes(x = lex)) +
  geom_histogram(aes(y = after_stat(density)), color = "black", fill = "turquoise4") +
  geom_density(color = "red")
```

## 16.7 Density Plots — `geom_density()`

```{r, eval=FALSE}
# Basic density plot
gap %>%
  filter(year == 1935) %>%
  ggplot(aes(x = lex)) +
  geom_density(color = "turquoise4", fill = "turquoise4", alpha = 0.4)

# Overlapping density plots — color and fill by year (use factor for discrete colors)
gap %>%
  filter(year %in% c(1930, 1950)) %>%
  mutate(year = factor(year)) %>%           # must be a factor for discrete coloring
  ggplot(aes(x = lex, color = year, fill = year)) +
  geom_density(alpha = 0.3) +
  labs(title = "Distribution of Life Expectancy, 1930 and 1950",
       x     = "Life Expectancy (years)",
       y     = NULL,
       color = "Year",
       fill  = "Year")
```

## 16.8 Labels and Titles — `labs()`

```{r, eval=FALSE}
# Add title, x label, y label, and legend titles
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = flipper_length_mm)) +
  geom_point(mapping = aes(color = species, shape = species)) +
  geom_smooth(method = lm, se = T) +
  labs(title = "Body Mass (g) vs. Flipper Length (mm)",
       x     = "Body Mass (g)",
       y     = "Flipper Length (mm)")

# Remove an axis label by setting it to NULL
homerange %>%
  ggplot(aes(x = taxon)) +
  geom_bar() +
  labs(title = "Observations by Taxon",
       x     = "Taxonomic Group",
       y     = NULL)   # removes the y-axis label entirely
```

## 16.9 Themes and Text Customization

```{r, eval=FALSE}
# theme() adjusts plot elements
# rel() scales text relative to the base size
# hjust controls horizontal alignment (0 = left, 0.5 = center, 1 = right)

life_history %>%
  ggplot(aes(x = mass, y = litter_size)) +
  geom_point(na.rm = T) +
  scale_x_log10() +
  geom_smooth(method = lm, se = F, na.rm = T) +
  labs(title = "Mass vs. Litter Size",
       x     = "Mass (g)",
       y     = "Litter Size") +
  theme(plot.title = element_text(size = rel(2), hjust = 0.5))

# Rotate x-axis text — useful for long category names
homerange %>%
  ggplot(aes(x = taxon, fill = trophic_guild)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Move legend position
p + theme_linedraw() +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 60, hjust = 1))

# Remove the legend
income_cat2000 %>%
  ggplot(aes(x = income_category, fill = income_category)) +
  geom_bar() +
  theme(legend.position = "none")

# Built-in themes
p + theme_gray()
p + theme_classic()
p + theme_minimal()
p + theme_linedraw()
p + theme_bw()
```

## 16.10 ggthemes

```{r, eval=FALSE}
# install.packages("ggthemes")
library(ggthemes)

# List all available ggthemes
ls("package:ggthemes")[grepl("theme_", ls("package:ggthemes"))]

# Apply a ggtheme
p + theme_wsj()     # Wall Street Journal style
q + theme_stata()   # Stata style
```

## 16.11 Colors — RColorBrewer and paletteer

```{r, eval=FALSE}
library(RColorBrewer)
library(paletteer)

# View a color palette (qualitative = for categorical, sequential = for ordered)
display.brewer.pal(8, "Accent")   # shows the palette with 8 colors

# Apply a brewer palette
# scale_fill_brewer()   = for fill aesthetic (bars, boxes)
# scale_colour_brewer() = for color aesthetic (points, lines)

p + scale_fill_brewer(palette = "Accent") +
  theme(legend.position = "bottom")

q + theme_classic() + scale_colour_brewer(palette = "Accent")

# paletteer — access hundreds of palettes
colors <- paletteer::palettes_d_names   # browse available palettes

my_palette <- paletteer_d("ggsci::uniform_startrek")
my_palette

# Preview it as a barplot
barplot(rep(1, 7), axes = FALSE, col = my_palette)

# Apply paletteer palette using scale_fill_manual()
p + scale_fill_manual(values = my_palette)

# Custom hex code colors
library(scales)
my_cols <- c("#66CDAA", "#FFCD00FF")
show_col(my_cols)   # preview in the plot panel

p + scale_fill_manual(values = my_cols)
q + scale_colour_manual(values = my_cols)
```

## 16.12 Faceting — Multi-Panel Plots

```{r, eval=FALSE}
# facet_wrap() — ribbon of panels by one variable
# nrow controls how many rows of panels
homerange %>%
  ggplot(aes(x = taxon, y = log10_mass, fill = thermoregulation)) +
  geom_boxplot() +
  facet_wrap(~ trophic_guild,
             nrow       = 2,
             labeller   = label_both) +           # labels include variable name
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "none",
        strip.text = element_text(size = 12, face = "bold"))

# facet_wrap() with density — compare distributions
homerange %>%
  ggplot(aes(x = log10_mass)) +
  geom_density(fill = "steelblue", alpha = 0.3) +
  facet_wrap(~ thermoregulation, ncol = 2)

# facet_grid() — rows ~ columns layout
# Use . to indicate no variable for that axis
homerange %>%
  ggplot(aes(x = taxon, y = log10_mass, fill = thermoregulation)) +
  geom_boxplot() +
  facet_grid(. ~ thermoregulation, labeller = label_both) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "none")

# facet_grid() with two variables (rows ~ columns)
homerange %>%
  ggplot(aes(x = taxon, y = log10_mass, fill = thermoregulation)) +
  geom_boxplot() +
  facet_grid(trophic_guild ~ thermoregulation, labeller = label_both) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "none",
        strip.text = element_text(size = 8, face = "bold"))
```

## 16.13 Pipe Directly into ggplot

```{r, eval=FALSE}
# You can pipe a data frame directly into ggplot without specifying data=
# The piped object becomes the data argument automatically

homerange %>%
  select(class, log10_hra) %>%
  filter(class == "mammalia" | class == "aves") %>%
  ggplot(aes(x = class, y = log10_hra)) +
  geom_boxplot() +
  labs(title = "Homerange Size of Birds vs Mammals")

# Use geom_col() after computing a summary
sharks %>%
  group_by(county) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = reorder(county, n), y = n)) +
  geom_col(fill = "#0099f9", alpha = 0.8) +
  labs(title = "Shark Incidents by County (1950-2022)",
       x     = NULL,
       y     = "n") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        plot.title  = element_text(size = 14, face = "bold")) +
  geom_text(aes(label = n), vjust = -0.2, size = 3, color = "black")  # add data labels
```

---

# 17. Mapping — `ggmap` and `leaflet`

## 17.1 Static Maps with `ggmap`

```{r, eval=FALSE}
library(ggmap)

# Register your Stadia Maps API key (free at stadiamaps.com)
register_stadiamaps("YOUR API KEY HERE", write = FALSE)

# Step 1: get the range of your coordinate data
spiders %>%
  select(latitude, longitude) %>%
  summary()

# Step 2: create the bounding box
# f = 0.03 adds 3% padding on all sides
lat  <- c(34.67, 41.80)
long <- c(-124.1, -115.5)
bbox <- make_bbox(long, lat, f = 0.03)

# Step 3: download the base map tile
# maptype options: "stamen_terrain", "stamen_toner", "stamen_watercolor", etc.
map1 <- get_stadiamap(bbox, maptype = "stamen_terrain", zoom = 7)

# Step 4: display the base map
ggmap(map1)

# Step 5: add data points on top of the map
ggmap(map1) +
  geom_point(data    = spiders,
             aes(longitude, latitude),
             size    = 1,
             color   = "red",
             alpha   = 0.8) +
  labs(x     = "Longitude",
       y     = "Latitude",
       title = "CA Usofila Records")
```

## 17.2 Interactive Maps with `leaflet`

```{r, eval=FALSE}
library(leaflet)

# Get min/max coordinates from data
spiders %>%
  summarize(lon_min = min(longitude, na.rm = T),
            lon_max = max(longitude, na.rm = T),
            lat_min = min(latitude,  na.rm = T),
            lat_max = max(latitude,  na.rm = T))

# Store bounding box values
lon_min <- -124.13
lon_max <- -115.49
lat_min <- 34.67
lat_max <- 41.8

# Build interactive map with Stadia Maps tiles
leaflet(spiders) %>%
  addProviderTiles(providers$Stadia.StamenTerrain) %>%
  addCircleMarkers(
    lng         = ~longitude,    # ~ means "use this column"
    lat         = ~latitude,
    radius      = 2,
    stroke      = FALSE,
    fillOpacity = 0.7
  ) %>%
  addScaleBar(position = "bottomleft") %>%
  fitBounds(lng1 = lon_min, lat1 = lat_min,
            lng2 = lon_max, lat2 = lat_max)

# Build interactive map with OpenStreetMap tiles (default, no API key needed)
leaflet(spiders) %>%
  addTiles() %>%
  addCircleMarkers(
    lng         = ~longitude,
    lat         = ~latitude,
    radius      = 2,
    stroke      = FALSE,
    fillOpacity = 0.7
  ) %>%
  addScaleBar(position = "bottomleft") %>%
  fitBounds(lng1 = lon_min, lat1 = lat_min,
            lng2 = lon_max, lat2 = lat_max)

# Custom marker color
leaflet(sharks_fatal) %>%
  addProviderTiles(providers$Stadia.StamenTerrain) %>%
  addCircleMarkers(
    lng         = ~longitude,
    lat         = ~latitude,
    radius      = 2,
    stroke      = FALSE,
    fillColor   = "red",         # custom fill color
    fillOpacity = 1
  ) %>%
  addScaleBar(position = "bottomleft") %>%
  fitBounds(lng1 = lon_min, lat1 = lat_min,
            lng2 = lon_max, lat2 = lat_max)
```

---

# 18. Shiny Apps

## 18.1 Basic Shiny Skeleton

```{r, eval=FALSE}
library(shiny)

# Minimum skeleton — type "shinyapp" + Shift + Tab in RStudio for snippet
ui <- fluidPage(
  # user interface elements go here
)

server <- function(input, output, session) {
  # reactive computations and outputs go here
}

shinyApp(ui, server)
```

## 18.2 selectInput() — Drop-Down Menu

```{r, eval=FALSE}
library(shiny)

# Single selectInput for x variable
ui <- fluidPage(
  selectInput("x",                           # id used in server to access this input
              "Select X Variable",           # label shown to user
              choices  = c("bill_length_mm",
                           "bill_depth_mm",
                           "flipper_length_mm",
                           "body_mass_g"),
              selected = "bill_length_mm")   # default selection
)

server <- function(input, output, session) { }
shinyApp(ui, server)
```

## 18.3 Complete Scatter Plot App (selectInput + renderPlot)

```{r, eval=FALSE}
library(shiny)
library(tidyverse)
library(palmerpenguins)

ui <- fluidPage(
  selectInput("x",
              "Select X Variable",
              choices  = c("bill_length_mm", "bill_depth_mm",
                           "flipper_length_mm", "body_mass_g"),
              selected = "bill_length_mm"),
  selectInput("y",
              "Select Y Variable",
              choices  = c("bill_length_mm", "bill_depth_mm",
                           "flipper_length_mm", "body_mass_g"),
              selected = "bill_depth_mm"),
  checkboxInput("show_lm",
                "Show best-fit line (lm)",
                value = FALSE),
  plotOutput("plot", width = "500px", height = "400px")
)

server <- function(input, output, session) {
  output$plot <- renderPlot({
    ggplot(penguins,
           aes(x = .data[[input$x]],   # .data[[]] accesses column by string name
               y = .data[[input$y]])) +
      geom_point(na.rm = T) +
      if (input$show_lm)
        geom_smooth(method = "lm", se = F, na.rm = T)
    else
      NULL
  })
}

shinyApp(ui, server)
```

## 18.4 radioButtons() Input

```{r, eval=FALSE}
library(shiny)

ui <- fluidPage(
  radioButtons("x",
               "Select Fill Variable",
               choices  = c("trophic_guild", "thermoregulation"),
               selected = "trophic_guild"),
  plotOutput("plot")
)

server <- function(input, output, session) {
  output$plot <- renderPlot({
    ggplot(data = homerange,
           aes_string(x = "locomotion", fill = input$x)) +
      geom_bar(position = "dodge", alpha = 0.8, color = "black") +
      labs(x = NULL, y = NULL, fill = "Fill Variable")
  })
}

shinyApp(ui, server)
```

## 18.5 shinythemes

```{r, eval=FALSE}
library(shinythemes)

# Add theme = shinythemes::shinytheme("themename") inside fluidPage()
# Available themes: "cyborg", "cerulean", "flatly", "darkly", etc.
ui <- fluidPage(
  theme = shinythemes::shinytheme("cyborg"),
  selectInput("y", "Select Sleep Variable",
              choices  = c("sleep_total", "sleep_rem", "sleep_cycle", "awake"),
              selected = "sleep_total"),
  plotOutput("plot", width = "600px", height = "500px")
)

server <- function(input, output, session) {
  output$plot <- renderPlot({
    msleep %>%
      filter(vore != "NA") %>%
      ggplot(aes(x     = vore,
                 y     = .data[[input$y]],
                 fill  = vore)) +
      geom_boxplot(alpha = 0.75) +
      labs(title = "Sleep Variables by Vore Type",
           x     = "Vore",
           fill  = "Vore Type") +
      theme_minimal()
  })
}

shinyApp(ui, server)
```

## 18.6 shinydashboard

```{r, eval=FALSE}
library(shiny)
library(shinydashboard)
library(tidyverse)

# shinydashboard replaces fluidPage() with dashboardPage()
# Minimum structure: header + sidebar + body

# Skeleton dashboard
ui <- dashboardPage(
  dashboardHeader(title = "My Dashboard"),
  dashboardSidebar(disable = T),   # disable = T hides the sidebar
  dashboardBody()
)
server <- function(input, output) { }
shinyApp(ui, server)

# ---- Full example with sidebar inputs ----
ui <- dashboardPage(
  dashboardHeader(title = "Distribution of Gestation Length by Order"),
  dashboardSidebar(
    selectInput("x",
                "Select Order",
                choices  = sort(unique(gestation$order)),   # sort() for alphabetical
                selected = "Afrosoricida"),
    selectInput("y",
                "Select Variable",
                choices  = c("bodymass", "longevity", "gestationlength"),
                selected = "gestationlength")
  ),
  dashboardBody(
    plotOutput("plot", width = "500px", height = "400px")
  )
)

server <- function(input, output, session) {
  output$plot <- renderPlot({
    gestation %>%
      filter(input$x == .data[["order"]]) %>%
      ggplot(aes(.data[[input$y]])) +
      geom_density(fill  = "steelblue",
                   color = "black",
                   alpha = 0.6) +
      theme_light()
  })
}

shinyApp(ui, server)

# ---- Wolf pack example ----
ui <- dashboardPage(
  dashboardHeader(title = "Sex Composition by Wolf Population"),
  dashboardSidebar(disable = T),
  dashboardBody(
    selectInput("x",
                "Select Population",
                choices  = sort(unique(wolves$pop)),
                selected = "AK.PEN"),
    plotOutput("plot", width = "500px", height = "400px")
  )
)

server <- function(input, output, session) {
  output$plot <- renderPlot({
    wolves %>%
      filter(pop == input$x) %>%
      ggplot(aes(x = sex, fill = sex)) +
      geom_bar(alpha = 0.8, color = "black") +
      labs(x = "Sex", y = NULL) +
      theme_minimal()
  })
}

shinyApp(ui, server)
```

---

# 19. Common Utility Functions & Patterns

## 19.1 Useful Standalone Functions

```{r, eval=FALSE}
# Round to n significant digits
round(pi / 2, 4)

# Base R histogram (quick check)
hist(life_history_no_nas$max_life)

# Store a plot as an object (then add layers or themes later)
p <- homerange %>%
  ggplot(aes(x = log10_mass, y = log10_hra))

p + geom_point()   # now add a geom
p + geom_jitter(size = 3)

# str_remove() from stringr — remove pattern from string
heartrate2 %>%
  mutate(patient = str_remove(patient, "_"))

# str_remove() on time strings (shark data)
sharks %>%
  mutate(time = str_remove(time, ":"))

# convert as_factor() — turns a column into a factor (preserving order/type)
sharks %>%
  ggplot(aes(x = as_factor(month), y = total)) +
  geom_col()
```

## 19.2 Exploratory Analysis Workflow

```{r, eval=FALSE}
# Recommended steps when you first load data:
data <- read_csv("file.csv") %>% clean_names()

dim(data)       # rows x columns
names(data)     # column names
glimpse(data)   # structure, types, preview
summary(data)   # summary stats, NA counts

# Check for NAs
data %>%
  summarize(across(everything(), ~ sum(is.na(.))))

# Or with naniar
miss_var_summary(data)

# Check factor levels after converting
data$col <- as.factor(data$col)
levels(data$col)

# Check for duplicate rows
data %>%
  count(col1, col2) %>%
  filter(n > 1)
```

## 19.3 Advanced Gapminder Analysis Patterns

```{r, eval=FALSE}
# Population growth between two years using pivot_wider
pop_growth_top5 <- gapminder %>%
  filter(year %in% c(1925, 2025)) %>%
  select(country, year, pop) %>%
  pivot_wider(names_from  = year,
              values_from = pop,
              names_prefix = "year_") %>%
  mutate(pop_growth     = year_2025 - year_1925,
         abs_pop_growth = abs(pop_growth)) %>%
  arrange(desc(abs_pop_growth)) %>%
  slice_head(n = 5)

# Global trends over time
global_yearly <- gapminder %>%
  group_by(year) %>%
  summarize(
    global_population    = sum(pop, na.rm = TRUE),
    weighted_mean_lex    = weighted.mean(lex, pop, na.rm = TRUE),  # population-weighted
    median_gdp           = median(gdp, na.rm = TRUE),
    .groups              = "drop"
  )

# Correlation between two variables by year
correlation_by_year <- gapminder %>%
  filter(gdp > 0) %>%
  group_by(year) %>%
  summarize(
    cor_log10gdp_lex = cor(log10(gdp), lex, use = "complete.obs"),
    .groups          = "drop"
  )

# Outlier detection using the IQR rule
outliers_2025 <- gapminder %>%
  filter(year == 2025) %>%
  select(country, pop, gdp, lex) %>%
  pivot_longer(cols = c(pop, gdp, lex), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  mutate(
    q1         = quantile(value, 0.25, na.rm = TRUE),
    q3         = quantile(value, 0.75, na.rm = TRUE),
    iqr        = q3 - q1,
    lower      = q1 - 1.5 * iqr,
    upper      = q3 + 1.5 * iqr,
    is_outlier = value < lower | value > upper
  ) %>%
  ungroup() %>%
  filter(is_outlier) %>%
  arrange(variable, desc(value)) %>%
  select(variable, country, value, lower, upper)
```

---

# 20. Package Quick Reference

```{r, eval=FALSE}
# ---- Load all commonly used packages ----
library("tidyverse")        # includes dplyr, ggplot2, tidyr, readr, stringr, etc.
library("janitor")          # clean_names(), tabyl()
library("naniar")           # miss_var_summary(), gg_miss_var(), replace_with_na()
library("palmerpenguins")   # penguins dataset
library("lubridate")        # date/time manipulation
library("RColorBrewer")     # color palettes for ggplot
library("paletteer")        # huge collection of palettes
library("ggthemes")         # extra ggplot themes
library("shiny")            # interactive web apps
library("shinythemes")      # themes for shiny
library("shinydashboard")   # dashboard UI for shiny
library("ggmap")            # static maps from Stadia/Google
library("leaflet")          # interactive maps
library("scales")           # show_col(), percent formatting
```

---

*This cheat sheet was compiled from a full semester of BIS 15L / Data Science in R coursework. All code blocks are knit-ready and can be run directly in RMarkdown after loading the appropriate packages and data files.*
